Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-11-13 docs/submission-mee-2020-09-02.tex
last mee version

email: \href{mailto:ejmctavish@gmail.com}{\nolinkurl{ejmctavish@gmail.com}}

@@

\textbf{Correspondence email}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:ejmctavish@gmail.com}{\nolinkurl{ejmctavish@gmail.com}}

@@

\textbf{Word count}: 2419

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automate
  some parts of the process of phylogenetic reconstruction, mainly molecular character matrix assembly,
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and choice of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is a command-line Python program that automates the update of published
  phylogenies by adding public DNA sequences to underlying alignments of previously published phylogenies. It also provides a framework for straightforward comparison of published phylogenies with their updated versions, by leveraging upon tools from the Open Tree of Life project to link taxonomic information across databases.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on publicly available expert phylogenetic knowledge.
  Phylogeneticists and taxonomic group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  The Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging more researchers to strive for better sharing practices. Physcraper can be used with any OS and is released under an open-source license. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenies capture the shared history of organisms and provide key evolutionary context for our biological observations.
Public biological databases constitute an amazing resource for evolutionary studies, but a large portion of public molecular data has never been incorporated into any phylogenetic estimate. Extending existing phylogenies with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we introduce Physcraper, a tool to build upon DNA datasets that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope, by using data from public DNA databases to update a starting tree and single locus alignments.

Taxonomic idiosyncrasies across databases represent a huge challenge for automatic integration of data into phylogenies; challenge that can be addressed with a unified taxonomy for name standardization. The Open Tree of Life project (OpenTree)
constructs a comprehensive tree of life by synthesizing published phylogenies along with taxonomic data. OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed in the tree based on taxonomy. To achieve this, OpenTree unifies taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), including the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}), among others. This OpenTree taxonomy represents a key resource for connecting data from any biological database that has been integrated to it.

Another challenhe for incorporating public molecular data into existing phylogenies is assembling high-quality homology hypotheses.
While genomics has, and will continue to, revolutionize phylogenetic inference, the variety of alternative genomic sequencing approaches produce largely non-overlapping genomic datasets across taxa, creating challenges in wide scale phylogenetic reconstruction.
Phylogenomics ameliorate this problem by focusing on targeted capture of informative characters from independent and single-copy genetic markers (Andermann \emph{et al.} \protect\hyperlink{ref-andermann2020guide}{2020}). Yet, decades of single locus sequencing have generated massive amounts of homologous DNA datasets that can be used for phylogenetic reconstruction at many scales. Moreover, species tree reconstructions from multiple single locus datasets using the multispecies coalescent model are considered the gold standard for inferring species relationships (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}).

More than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of 98.05\% of the almost 241,000 distinct taxa in the NCBI taxonomy at the time (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it is a largely time consuming and mostly non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to infer phylogenetic relationships in a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
However, thoughtfully curated markers and alignments can improve phylogenetic reconstructions, even in phylogenomic reconstructions (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the benefits from massive amounts of newly released molecular data and fine-grained curation from human experts, is to rely on manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8,200 alignments from published peer-reviewed studies, providing information on evolutionary relationships of around 100,000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}. Linking published alignments generated by experts with molecular data that has not yet been included in any public phylogenetic estimate has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life.

Physcraper relies on programmatic access protocols (API's) and the OpenTree taxonomy to automatically link phylogenetic data to molecular data from GenBank, alignments from TreeBASE, and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various OpenTree tools for comparison of existing phylogenetic hypotheses with newly generated ones.
Physcraper is coded as a Python pipeline that focuses on data interoperability, by using the OpenTree taxonomy as a way to link taxon data from different databases. This integration also allows users to rapidly place new data from a diverse range of biological databases in an evolutionary context, opening the possibility for a variety of comparative downstream analyses.

@@

{\centering \includegraphics[width=0.85\linewidth]{docs/figs/schematic-3}

@@

The general Physcraper framework is depicted in Figure \ref{fig:framework}. It consists of 4 steps: 1) identifying and processing a tree and its underlying alignment; 2) performing a BLAST search of DNA sequences from the original alignment on GenBank, and filtering of new sequences; 3) profile-aligning new sequences to the original alignment; 4) performing a phylogenetic analysis and comparing the updated tree to previous phylogenetic estimates.

@@

Taxon names in the input tree must be standardized to the OpenTree taxonomy (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) using OpenTree's bulk Taxonomic Name Resolution Service \href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool. Users can upload their own tree, or choose from among the 2, 950 standardized trees stored in OpenTree's \href{https://github.com/opentreeoflife/phylesystem}{Phylesystem} that also have alignments avilable on TreeBASE (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}).

The input alignment is a single locus alignment that was used in part or in whole to generate the input tree. If it is stored in TreeBASE, Physcraper retrieves it automatically. Otherwise, the user must provide the path to a local copy of the alignment.
Only taxa that are both in the sequence alignment and in the tree are considered further for analysis; at least one taxon and its corresponding sequence are required.

@@

The Basic Local Alignment Search Tool, BLAST (Altschul \emph{et al.} \protect\hyperlink{ref-altschul1990basic}{1990}) is used for DNA sequence search on a remote or local GenBank database. It is constrained to a ``search taxon'', a taxonomic group in the NCBI taxonomy that is automatically identified using the OpenTree \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{API} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), as the most recent common ancestor of the ingroup taxa in the tree that is also a named clade in the NCBI taxonomy (Fig. \ref{fig:framework}).
Alternatively, users can arbitrarily define a search taxon that is either a more or less inclusive clade relative to the ingroup of the input tree.

BLAST is implemented with the BLAST command line tools \texttt{blastn} function (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) and the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function from the \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excuded for analysis if 1) they do not belong to the search taxon; 2) they have an e-value above the cutoff (default to 0.00001); 3) they fall outside a min and max length threshold, defined as the proportion of the average length without gaps of all sequences in the input alignment (default values of 80\% and 120\%, respectively); 4) or if they are either identical to or shorter than an existing sequence in the input alignment and they represent the same taxon in the OpenTree taxonomy or the NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed (default to 5).

@@

A Maximum Likelihood (ML) gene tree is reconstructed for each alignment provided, using RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014}) with default settings (GTRCAT model and 100 bootstrap replicates with the default algorithm) and using the input tree as a starting tree for the ML searches.
Bootstrap results are summarized with DendroPy's SumTrees module (current version 4.4.0; Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
The updated tree is compared to the original tree using an automated conflict analysis which calculates Robinson Foulds weighted and unweighted metrics using Dendropy (Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}), and performs a node by node comparison between the synthetic OpenTree and the original and updated tree individually, using OpenTree's conflict API (Redelings \& Holder \protect\hyperlink{ref-redelings2017supertree}{2017}).
For the conflict analysis to be meaningful, the root of the tree needs to be accurately identified. The best way to currenty do this is by hand, although a default rooting based on OpenTree's taxonomy is also available.

@@

A user is interested in the phylogenetic relationships within the genus \emph{Ilex}. Commonly known as ``hollies'', the genus encompasses between 400-700 living species, and is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.

An online literature review in June 2020 (google scholar search for ``ilex phylogeny'') reveals that there are several published phylogenetic trees showing relationships within the hollies (Cuénoud \emph{et al.} \protect\hyperlink{ref-cuenoud2000molecular}{2000}; Setoguchi \& Watanabe \protect\hyperlink{ref-setoguchi2000intersectional}{2000}; Selbach-Schnadelbach \emph{et al.} \protect\hyperlink{ref-selbach2009new}{2009}; Manen \emph{et al.} \protect\hyperlink{ref-manen2010history}{2010}), but only two have their data available publicly (Gottlieb \emph{et al.} \protect\hyperlink{ref-gottlieb2005molecular}{2005}; Yao \emph{et al.} \protect\hyperlink{ref-yao2020phylogeny}{2020}).
Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}) made original tree and alignment data available in \href{https://treebase.org/treebase-web/search/study/summary.html?id=1091}{TreeBASE}. The ``Gottlieb2005'' tree sampling 41 species was added to the \href{https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home}{OpenTree Phylesystem} and its information has been integrated into \href{https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa}{OpenTree's synthetic tree}.

The most recent \emph{Ilex} tree from Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}), has been made available in the \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. The ``Yao2020'' tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.

Figure \ref{fig:results} shows results from the Physcraper analysis of an alignment of the internal transcribed spacer DNA region (ITS) from Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}). Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional \emph{Ilex} taxa. The best RaxML tree is 99\% resolved, with 25\% of nodes with bootstrap support \textless{} 0.1 and 48\% nodes with bootstrap support \textgreater{} 0.75.
A large portion of internal branches are negligibly small, with 30 branches \textless{} 0.00001 substitution rate units, from which only 9 have a bootstrap support \textgreater{} 0.75 (Fig. \ref{fig:results}).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
and contributes phylogenetic data to 134 additional \emph{Ilex} taxa, from which
67 are also in the Physcraper updated Gottlieb2005 tree. While Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}) also used ITS as a marker, their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis. However, Physcraper was able to incorporate 18 taxa that were not in Yao2020.
This might be caused by the method they used to download existing ITS \emph{Ilex} sequences from GenBank, which is not fully explained in the publication, but seems to be a ``manual'' process.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-results} \caption{A) Phylogenetic tree obtained by updating the Gottlieb et al. 2005 tree in B) using Physcraper.}\label{fig:results}
\end{figure}

@@

\emph{Figure \ref{fig:results} caption continued}: Tips in original alignment and new tips added with Physcraper are depicted in black and red, respectively. Physcraper obtained sequences from the GenBank database via local BLAST of all sequences in the original alignment that generated tree in B), filtered them following criteria specified in section ``DNA sequence search and filtering'', aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML with 100 bootstraps. B-D show results of the conflict analysis performed with OpenTree tools.

@@

Databases that preserve and democratize access to biological data
are essential resources for our scientific endeavor.
New molecular data keep accumulating and tools facilitating its integration into existent evolutionary knowledge are essential.

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of public molecular data (e.g., Phylota (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017})) focus on generating full phylogenies \emph{de novo}, i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by summarizing previous phylogenetic estimates.
While Physcraper does not generate phylogenies \emph{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and accelerating the enrichment and updating of phylogenetic relationships in any region of the tree of life.
The PUMPER pipeline (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}) also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to benchmark a comparison.

Physcraper generates individual gene trees, which fail to capture the complexity of species' evolutionary history (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}). Yet, Physcraper makes it straightforward to gather alignments and gene trees for multiple loci from a taxonomic group of interest, which can then be used to perform coalescent analysis to generate species trees with e.g., ASTRAL (Mirarab \emph{et al.} \protect\hyperlink{ref-mirarab2014astral}{2014}), BEAST2 (Bouckaert \emph{et al.} \protect\hyperlink{ref-bouckaert2019beast}{2019}), or SVD Quartets (Chifman \& Kubatko \protect\hyperlink{ref-chifman2014quartet}{2014})).

Physcraper has the potential to link phylogenies to data available in any of the taxonomies integrated by the OpenTree taxonomy (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), such as geographical locations from the Global Biodiversity Information Facility, or fossils from the Paleobiology Database.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically (Webb \emph{et al.} \protect\hyperlink{ref-webb2010biodiversity}{2010}),
obtaining trees for ecophylogenetic studies (Helmus \& Ives \protect\hyperlink{ref-helmus2012phylogenetic}{2012}),
systematizing molecular (and other) databases, i.e., curating taxonomic assignations (San Mauro \& Agorreta \protect\hyperlink{ref-san2010molecular}{2010}),
and generating custom trees for ecological and evolutionary downstream analyses (Stoltzfus \emph{et al.} \protect\hyperlink{ref-stoltzfus2013phylotastic}{2013}).

>>>>

email: \href{mailto:ejmctavish@ucmerced.edu}{\nolinkurl{ejmctavish@ucmerced.edu}}

@@

\textbf{Correspondence email}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:ejmctavish@ucmerced.edu}{\nolinkurl{ejmctavish@ucmerced.edu}}

@@

\textbf{Word count}: 2969

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automate
  some parts of the process of phylogenetic reconstruction, mainly molecular character matrix assembly,
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and selection of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is a command-line Python program that automates the update of published
  phylogenies by adding public DNA sequences to underlying alignments of previously published phylogenies. It also provides a framework for straightforward comparison of published phylogenies with their updated versions, by leveraging upon tools from the Open Tree of Life project to link taxonomic information across databases.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on publicly available expert phylogenetic knowledge.
  Phylogeneticists and taxonomic group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  The Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging researchers to strive for better sharing practices. Physcraper can be used with any OS and is released under an open-source license. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenies capture the shared history of organisms and provide key evolutionary context for our biological observations.
Public biological databases constitute an amazing resource for evolutionary studies. Updating existing phylogenies with molecular data that has never been incorporated into any phylogenetic estimate, geographical location, fossils, and other data in a reproducible and continuous manner is possible by establishing a data interoperability framework for biological databases. Here, we introduce Physcraper, a tool that automates database connections to build upon homology hypotheses that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope to update a starting tree and single locus alignments with public DNA data.

Taxonomic idiosyncrasies across databases represent a huge challenge for automatic integration of data into phylogenies, which can be addressed with a unified taxonomy for name standardization. The Open Tree of Life project (OpenTree)
constructs a comprehensive tree of life by synthesizing published phylogenies and taxonomy. OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed in the tree based on taxonomy. To achieve this, OpenTree unifies taxonomic data from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), including the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}), among others. The OpenTree taxonomy represents a key resource for connecting data from any biological database that has been integrated to it.

Another challenge for incorporating public molecular data into existing phylogenies is assembling high-quality homology hypotheses.
While genomics has, and will continue to, revolutionize phylogenetic inference, the variety of alternative genomic sequencing approaches it uses produce largely non-overlapping genomic datasets across taxa, creating challenges in wide scale phylogenetic reconstruction.
Phylogenomics ameliorate this problem by focusing on targeted capture of informative loci (Andermann \emph{et al.} \protect\hyperlink{ref-andermann2020guide}{2020}). Yet, decades of single locus sequencing have generated massive amounts of homologous DNA datasets that can be used for phylogenetic reconstruction at many scales.

More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72 million DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of 98.05\% of the almost 241,000 distinct taxa in the NCBI taxonomy at the time (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it is a largely time consuming and mostly non-reproducible approach. Computational pipelines that mine DNA databases fast, efficiently, and reproducibly, have been applied to infer phylogenetic relationships in a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
However, fine-grained curated markers and alignments can improve phylogenetic reconstructions, even in phylogenomic analyses (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

There are almost 8,200 publicly available, peer-reviewed alignments, covering around 100,000 distinct taxa in the TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}), which can be used as seeds to mine molecular databases, and as ``jump-start'' alignments for phylogenetic reconstructions (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}) to continually enrich, update and compare existing phylogenetic knowledge.

Physcraper is a Python pipeline using OpenTree's taxonomy and programmatic access protocols (API's) to implement a database interoperability framework that automatically links phylogenies that have been standardized to OpenTree taxonomy, to alignments from TreeBASE, data from GenBank, and phylogenies from OpenTree's Phylesystem.
Physcraper aims to demonstrate the benefits of reproducible workflows and open science in phylogenetics, and encourage better data sharing practices in the community.

@@

{\centering \includegraphics[width=0.85\linewidth]{docs/figs/schematic-4}

@@

The general Physcraper framework consists of 4 steps (Fig. \ref{fig:framework}): 1) identifying and processing a tree and its underlying alignment; 2) performing a BLAST search of DNA sequences from original alignment on GenBank, and filtering of new sequences; 3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic analysis and comparing the updated tree to existing phylogenies.

@@

Taxon names in the input tree must be standardized to OpenTree taxonomy (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) using OpenTree's bulk Taxonomic Name Resolution Service \href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool. Users can upload their own tree, or choose from among the 2, 950 standardized trees stored in OpenTree's \href{https://github.com/opentreeoflife/phylesystem}{Phylesystem} that also have alignments available on TreeBASE (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}).

The input alignment is a single locus DNA dataset that was used in part or in whole to generate the input tree. Physcraper retrieves TreeBASE alignments automatically. Alternatively, users must provide the path to a local copy of the alignment.
Only taxa that are both in the sequence alignment and in the tree are considered further for analysis; at least one taxon and its corresponding sequence are required.

@@

The Basic Local Alignment Search Tool, BLAST (Altschul \emph{et al.} \protect\hyperlink{ref-altschul1990basic}{1990}) is used for DNA sequence search on a remote or local GenBank database. It is constrained to a ``search taxon'', a taxonomic group in the NCBI taxonomy that is automatically identified using the OpenTree \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{API} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), as the most recent common ancestor of ingroup taxa that is also a named clade in the NCBI taxonomy (Fig. \ref{fig:framework}).
Alternatively, users can arbitrarily define a search taxon that is either a more or less inclusive clade relative to the ingroup taxa.

BLAST is implemented with the \texttt{blastn} function (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) and the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function from \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} modified to accept an alternative BLAST address.
Each sequence in the alignment is BLASTed once against all DNA sequences in GenBank.
New sequences are excluded for analysis if they 1) are not in the search taxon; 2) have an e-value above the cutoff (default to 0.00001); 3) fall outside a min and max length threshold, defined as the proportion of the average length without gaps of all sequences in input alignment (default values of 80\% and 120\%, respectively); 4) or if they are either identical to or shorter than an existing sequence in the input alignment and they represent the same taxon in OpenTree or NCBI taxonomy.
An arbitrary maximum number of randomly chosen sequences per taxon are allowed (default to 5).

@@

RAxML (Stamatakis \protect\hyperlink{ref-stamatakis2014raxml}{2014}) is implemented to reconstruct a Maximum Likelihood (ML) gene tree for each input alignment with default settings (GTRCAT model and 100 bootstrap replicates with default algorithm), using input tree as starting tree for ML searches.
Bootstrap results are summarized using DendroPy's SumTrees module (Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}).

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original tree are compared with Robinson-Foulds weighted and unweighted metrics estimated with Dendropy (Sukumaran \& Holder \protect\hyperlink{ref-sukumaran2010dendropy}{2010}), and with a node by node comparison between the synthetic OpenTree and original and updated tree individually, using OpenTree's conflict API (Redelings \& Holder \protect\hyperlink{ref-redelings2017supertree}{2017}).

@@

A user is interested in phylogenetic relationships within the genus \emph{Ilex}. Commonly known as ``hollies'', the genus encompasses between 400-700 living species, and is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.

An online literature review in June 2020 (Google scholar search for ``ilex phylogeny'') reveals that there are several published phylogenies showing relationships within the hollies (Cuénoud \emph{et al.} \protect\hyperlink{ref-cuenoud2000molecular}{2000}; Setoguchi \& Watanabe \protect\hyperlink{ref-setoguchi2000intersectional}{2000}; Selbach-Schnadelbach \emph{et al.} \protect\hyperlink{ref-selbach2009new}{2009}; Manen \emph{et al.} \protect\hyperlink{ref-manen2010history}{2010}), but only two have data publicly available (Gottlieb \emph{et al.} \protect\hyperlink{ref-gottlieb2005molecular}{2005}; Yao \emph{et al.} \protect\hyperlink{ref-yao2020phylogeny}{2020}).
Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}) made original tree and alignment available in \href{https://treebase.org/treebase-web/search/study/summary.html?id=1091}{TreeBASE}. The ``Gottlieb2005'' tree sampling 41 species was added to \href{https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home}{OpenTree Phylesystem} and it has been integrated into \href{https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa}{OpenTree's synthetic tree}.

The most recent \emph{Ilex} tree (Yao \emph{et al.} \protect\hyperlink{ref-yao2020phylogeny}{2020}) is available in \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{OpenTree Phylesystem} and in the \href{https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x}{DRYAD repository}. With 175 tips, the ``Yao2020'' tree is the best sampled phylogeny yet available for the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed spacer DNA region (ITS) alignment from Gottlieb \emph{et al.} (\protect\hyperlink{ref-gottlieb2005molecular}{2005}), using a local GenBank database. BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap analyses taking an additional 13hrs.
The updated Gottlieb2005 tree (Fig. \ref{fig:results}) displays all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional \emph{Ilex} taxa. The best RaxML tree is 99\% resolved, with 25\% of nodes with bootstrap support \textless{} 0.1 and 48\% nodes with bootstrap support \textgreater{} 0.75.
A large portion of internal branches are negligibly small, with 30 branches \textless{} 0.00001 substitution rate units, from which only 9 have a bootstrap support \textgreater{} 0.75 (Fig. \ref{fig:results}).
For comparison, Yao2020 also contains all 41 distinct taxa from the original Gottlieb2005 study,
and contributes phylogenetic data to 134 additional \emph{Ilex} taxa, from which
67 are also in updated Gottlieb2005. While Yao \emph{et al.} (\protect\hyperlink{ref-yao2020phylogeny}{2020}) also used ITS as a marker, their GenBank data is not released yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis. However, Physcraper was able to incorporate 18 taxa that were not in Yao2020.

\begin{figure}
\includegraphics[width=1\linewidth]{docs/figs/ilex-results} \caption{A) Phylogeny updated with Physcraper from Gottlieb et al. 2005 tree in B.}\label{fig:results}
\end{figure}

@@

\emph{Figure \ref{fig:results} caption continued}: Tips in original alignment and new tips added with Physcraper are depicted in black and red, respectively. Physcraper obtained sequences from the GenBank database via local BLAST of all sequences in the original alignment that generated tree in B), filtered them following criteria from section ``DNA sequence search and filtering'', aligned them to original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML with 100 bootstraps. B-D conflict analyses performed with OpenTree tools.

@@

Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into existent evolutionary knowledge are needed.

Phylogenetic pipelines designed to make evolutionary sense of the vast amount of public molecular data (e.g., Phylota (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017})) focus on generating full phylogenies \emph{de novo}, i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by summarizing previous phylogenetic estimates.
While Physcraper does not generate phylogenies \emph{de novo} in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the importance of
open science in facilitating phylogenetic placement of public molecular data and accelerating enrichment and updating of phylogenetic relationships in any region of the tree of life.
The PUMPER pipeline (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}) also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation was unsuccessful following instructions from the author, and a comparison analysis was unfeasible.

Physcraper generates individual gene trees, failing to capture the complexity of species' evolutionary history (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}). Yet, Physcraper facilitates gathering alignments and gene trees for multiple loci from a group of interest, that can be used to reconstruct species trees with ASTRAL (Mirarab \emph{et al.} \protect\hyperlink{ref-mirarab2014astral}{2014}), BEAST2 (Bouckaert \emph{et al.} \protect\hyperlink{ref-bouckaert2019beast}{2019}), or SVD Quartets (Chifman \& Kubatko \protect\hyperlink{ref-chifman2014quartet}{2014})).

Physcraper can potentially link phylogenies to data available in any of the taxonomies integrated in the OpenTree taxonomy (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}), such as geographical locations from the Global Biodiversity Information Facility, or fossils from the Paleobiology Database.
The Physcraper workflow can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically (Webb \emph{et al.} \protect\hyperlink{ref-webb2010biodiversity}{2010}),
systematizing molecular (and other) databases, i.e., curating taxonomic assignations (San Mauro \& Agorreta \protect\hyperlink{ref-san2010molecular}{2010}),
and generating custom trees for ecological (Helmus \& Ives \protect\hyperlink{ref-helmus2012phylogenetic}{2012}) and evolutionary downstream analyses (Stoltzfus \emph{et al.} \protect\hyperlink{ref-stoltzfus2013phylotastic}{2013}).

Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.
Initial ideas about the data are constantly changed by results from new analyses.
Physcraper provides a framework for reproducible phylogenetics that has the potential to consistently provide context for these ideas, highlighting the importance of data sharing and open science in the field, biology and science.
